---
title: 'Lab 2: Community'
author: "Halina Do-Linh"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Clustering

## Load and plot `penguins` dataset

```{r}
# load R packages
librarian::shelf(
  dplyr, DT, ggplot2, palmerpenguins, skimr, tibble)

# set seed for reproducible results
set.seed(42)

# load the dataset
data("penguins")
```


```{r}
# look at documentation in RStudio
if (interactive())
  help(penguins)

# show data table
datatable(penguins)
```


```{r}
# skim the table for a summary
skim(penguins)
```


```{r}
# remove the rows with NAs
penguins <- na.omit(penguins)

# plot bill length vs width, species naive
ggplot(
  penguins, aes(bill_length_mm, bill_depth_mm)) +
  geom_point()
```


```{r}
# plot bill length vs width, color by species
legend_pos <- theme(
    legend.position = c(0.95, 0.05),
    legend.justification = c("right", "bottom"),
    legend.box.just = "right")
ggplot(
  penguins, aes(bill_length_mm, bill_depth_mm, color = species)) +
  geom_point() +
  legend_pos
```

## Cluster `penguins` using `kmeans()`

```{r}
# cluster using kmeans
k <- 3  # number of clusters
penguins_k <- kmeans(
  penguins %>% 
    select(bill_length_mm, bill_depth_mm), 
  centers = k)

# show cluster result
penguins_k
```


```{r}
# compare clusters with species (which were not used to cluster)
table(penguins_k$cluster, penguins$species)
```

## Question

How many observations could be considered “misclassified” if expecting bill length and width to differentiate between species?

**Answer:** At least 45 observations for the gentoo penguin could be misclassifed since most gentoo observations are in the third cluster. There could also be some misclassification in the third cluster where chinstrap and adelie penguins have been added to that cluster.

```{r}
# extract cluster assignment per observation
Cluster = factor(penguins_k$cluster)

ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color = Cluster)) +
  geom_point() + 
  legend_pos
```


## Question

Comparing the observed species plot with 3 species with the kmeans() cluster plot with 3 clusters, where does this “unsupervised” kmeans() technique (that does not use species to “fit” the model) produce similar versus different results? One or two sentences would suffice. Feel free to mention ranges of values along the axes.

**Answer:** The observed species plot with the unsupervised `kmeans()` technique visually shows the three species grouped as diagonal circles. The plot with the `kmeans()` technique visually shows the clusters grouped as horizontals. When comparing the two plots it appears the `kmeans()` clusters are mostly similar, but some of the gentoo and chinstrap observations have been predicted to be other species or misclassified.  


## Plot Voronoi diagram of clustered `penguins`

```{r}
librarian::shelf(ggvoronoi, scales)

# define bounding box for geom_voronoi()
xr <- extendrange(range(penguins$bill_length_mm), f=0.1)
yr <- extendrange(range(penguins$bill_depth_mm), f=0.1)
box <- tribble(
  ~bill_length_mm, ~bill_depth_mm, ~group,
  xr[1], yr[1], 1,
  xr[1], yr[2], 1,
  xr[2], yr[2], 1,
  xr[2], yr[1], 1,
  xr[1], yr[1], 1) %>% 
  data.frame()

# cluster using kmeans
k <- 3  # number of clusters
penguins_k <- kmeans(
  penguins %>% 
    select(bill_length_mm, bill_depth_mm), 
  centers = k)

# extract cluster assignment per observation
Cluster = factor(penguins_k$cluster)

# extract cluster centers
ctrs <- as.data.frame(penguins_k$centers) %>% 
  mutate(
    Cluster = factor(1:k))

# plot points with voronoi diagram showing nearest centroid
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color = Cluster)) +
  geom_point() + 
  legend_pos +
  geom_voronoi(
    data = ctrs, aes(fill=Cluster), color = NA, alpha=0.5, 
    outline = box) + 
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_point(
    data = ctrs, pch=23, cex=2, fill="black")
```

**Task:** Show the Voronoi diagram for fewer (k=2) and more (k=8) clusters to see how assignment to cluster centroids work.

```{r}
# voronoi diagram for k=2
librarian::shelf(ggvoronoi, scales)

# define bounding box for geom_voronoi()
xr <- extendrange(range(penguins$bill_length_mm), f=0.1)
yr <- extendrange(range(penguins$bill_depth_mm), f=0.1)
box <- tribble(
  ~bill_length_mm, ~bill_depth_mm, ~group,
  xr[1], yr[1], 1,
  xr[1], yr[2], 1,
  xr[2], yr[2], 1,
  xr[2], yr[1], 1,
  xr[1], yr[1], 1) %>% 
  data.frame()

# cluster using kmeans
k <- 2  # number of clusters
penguins_k <- kmeans(
  penguins %>% 
    select(bill_length_mm, bill_depth_mm), 
  centers = k)

# extract cluster assignment per observation
Cluster = factor(penguins_k$cluster)

# extract cluster centers
ctrs <- as.data.frame(penguins_k$centers) %>% 
  mutate(
    Cluster = factor(1:k))

# plot points with voronoi diagram showing nearest centroid
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color = Cluster)) +
  geom_point() + 
  legend_pos +
  geom_voronoi(
    data = ctrs, aes(fill=Cluster), color = NA, alpha=0.5, 
    outline = box) + 
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_point(
    data = ctrs, pch=23, cex=2, fill="black")
```


```{r}
# voronoi diagram for k=8
librarian::shelf(ggvoronoi, scales)

# define bounding box for geom_voronoi()
xr <- extendrange(range(penguins$bill_length_mm), f=0.1)
yr <- extendrange(range(penguins$bill_depth_mm), f=0.1)
box <- tribble(
  ~bill_length_mm, ~bill_depth_mm, ~group,
  xr[1], yr[1], 1,
  xr[1], yr[2], 1,
  xr[2], yr[2], 1,
  xr[2], yr[1], 1,
  xr[1], yr[1], 1) %>% 
  data.frame()

# cluster using kmeans
k <- 8  # number of clusters
penguins_k <- kmeans(
  penguins %>% 
    select(bill_length_mm, bill_depth_mm), 
  centers = k)

# extract cluster assignment per observation
Cluster = factor(penguins_k$cluster)

# extract cluster centers
ctrs <- as.data.frame(penguins_k$centers) %>% 
  mutate(
    Cluster = factor(1:k))

# plot points with voronoi diagram showing nearest centroid
ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color = Cluster)) +
  geom_point() + 
  legend_pos +
  geom_voronoi(
    data = ctrs, aes(fill=Cluster), color = NA, alpha=0.5, 
    outline = box) + 
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_point(
    data = ctrs, pch=23, cex=2, fill="black")
```

# Hierarchial Clustering

```{r}
librarian::shelf(
  cluster, vegan)

# load dune dataset from package vegan
data("dune")

# show documentation on dataset if interactive
if (interactive())
  help(dune)
```


### Question
What are the rows and columns composed of in the `dune` data frame?

**Answer:** The `dune` data frame are observations of 30 species at 20 sites. The columns represent names of the species, abbreviated to 4+4 letters where the first four letters of the species and the last four are BLANK. 

### Calculate Ecological Distances on `sites`

```{r}
sites <- tribble(
  ~site, ~sp1, ~sp2, ~sp3,
    "A",    1,    1,    0,
    "B",    5,    5,    0,
    "C",    0,    0,    1) %>% 
  column_to_rownames("site")
sites
```


```{r}
sites_manhattan <- vegdist(sites, method="manhattan")
sites_manhattan
```


```{r}
sites_euclidean <- vegdist(sites, method="euclidean")
sites_euclidean
```


```{r}
sites_bray <- vegdist(sites, method="bray")
sites_bray
```

## Question

In your own words, how does Bray Curtis differ from Euclidean distance? See sites_euclidean versus sites_bray from lab code, slides from Lecture 05. Clustering and reading Chapter 8 of Kindt and Coe (2005).

**Answer:** The Bray Curtis distance differs from the Euclidean distance because it restricts the outputs to a range of 0 to 1 (where 0 is most similar and 1 is least similar). Euclidean distance is unrestricted, and is more influenced by abundance and outliers.


### Agglomerative hierarchical clustering on `dune`

```{r}
# Dissimilarity matrix
d <- vegdist(dune, method="bray")
dim(d)
```


```{r}
as.matrix(d)[1:5, 1:5]
```


```{r}
# Hierarchical clustering using Complete Linkage
hc1 <- hclust(d, method = "complete" )

# Dendrogram plot of hc1
plot(hc1, cex = 0.6, hang = -1)
```

## Question

Which function comes first, vegdist() or hclust(), and why? See HOMLR 21.3.1 Agglomerative hierarchical clustering.

**Answer:** The function `vegdist()` comes first because it is measuring the dissimilarity values (aka distances) before it assigns a cluster. Then you use `hclust()` to create the clusters that make up the dendrogram. 


```{r}
# Compute agglomerative clustering with agnes
hc2 <- agnes(dune, method = "complete")

# Agglomerative coefficient
hc2$ac
```


```{r}
# Dendrogram plot of hc2
plot(hc2, which.plot = 2)
```


## Question

In your own words how does hclust() differ from agnes()? See HOMLR 21.3.1 Agglomerative hierarchical clustering and help documentation (?hclust(), ?agnes()).

**Answer:** The `agnes()` function creates a dendrogram from the bottom up, meaning it builds upon similar paired clusters. Where as `hclust()` joins the two most similar clusters until there is a single one. 

distance between observations 

Both methods start at the bottom up

```{r}
# methods to assess
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

# function to compute coefficient
ac <- function(x) {
  agnes(dune, method = x)$ac
}

# get agglomerative coefficient for each linkage method
purrr::map_dbl(m, ac)
```


```{r}
# Compute ward linkage clustering with agnes
hc3 <- agnes(dune, method = "ward")

# Agglomerative coefficient
hc3$ac
```


```{r}
# Dendrogram plot of hc3
plot(hc3, which.plot = 2)
```

### Divisive hierarchical clustering on `dune`

```{r}
# compute divisive hierarchical clustering
hc4 <- diana(dune)

# Divise coefficient; amount of clustering structure found
hc4$dc
```

### Determining optimal clusters

```{r}
librarian::shelf(factoextra)

# Plot cluster results
p1 <- fviz_nbclust(dune, FUN = hcut, method = "wss",  k.max = 10) +
  ggtitle("(A) Elbow method")

p2 <- fviz_nbclust(dune, FUN = hcut, method = "silhouette", k.max = 10) +
  ggtitle("(B) Silhouette method")

p3 <- fviz_nbclust(dune, FUN = hcut, method = "gap_stat", k.max = 10) +
  ggtitle("(C) Gap statistic")

# Display plots side by side
gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```

### Working with dendrograms
```{r}
# Construct dendorgram for the Ames housing example
hc5 <- hclust(d, method = "ward.D2" )
dend_plot <- fviz_dend(hc5)
```


```{r}
dend_data <- attr(dend_plot, "dendrogram")
dend_cuts <- cut(dend_data, h = 8)
fviz_dend(dend_cuts$lower[[2]])
```


```{r}
# Ward's method
hc5 <- hclust(d, method = "ward.D2" )

# Cut tree into 4 groups
k = 4
sub_grp <- cutree(hc5, k = k)

# Number of members in each cluster
table(sub_grp)
```


```{r}
# Plot full dendogram
fviz_dend(
  hc5,
  k = k,
  horiz = TRUE,
  rect = TRUE,
  rect_fill = TRUE,
  rect_border = "jco",
  k_colors = "jco")
```