---
title: 'Lab 2: Community Cluster'
author: "Halina Do-Linh"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

librarian::shelf(
  palmerpenguins,
  tidyverse,
  janitor,
  DT,
  tibble,
  ggvoronoi,
  scales,
  cluster,
  vegan,
  factoextra
)

# set seed for reproducible results
set.seed(42)
```

# Load data from Palmer Penguins

```{r}
penguins <- read_csv(path_to_file("penguins_raw.csv")) %>% 
  clean_names() %>%
  dplyr::select(c("species", "flipper_length_mm", "body_mass_g")) %>% 
  na.omit() # had to remove rows with na to run kmeans function

# show data table
datatable(penguins)
```

```{r}
# plot flipper length vs body mass 
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
         geom_point()
```

```{r}
# plot flipper length vs body mass, color by species
legend_pos <- theme(
    legend.position = c(1, 0.01),
    legend.justification = c("right", "bottom"),
    legend.box.just = "right")
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  legend_pos
```

# Clustering

## K-Means Clustering

```{r}
# clutser using k means
k <- 3  # number of clusters
penguins_k <- kmeans(
  penguins %>%
    dplyr::select(flipper_length_mm, body_mass_g), # specify dplyr select bc of R base conflicts
  centers = k)

# show cluster result
penguins_k
```

```{r}
# compare clusters with species (which were not used to cluster)
table(penguins_k$cluster, penguins$species)
```

## Plot Voronoi Diagram

```{r}
# define bounding box for geom_voronoi()
box <- tribble(
  ~flipper_length_mm, ~body_mass_g, ~group,
  1, 0.1, 1,
  1, 2.5, 1,
  7, 2.5, 1,
  7, 0.1, 1,
  1, 0.1, 1) %>% 
  data.frame()

# cluster using kmeans
k <- 3  # number of clusters
penguins_k <- kmeans(
  penguins %>% 
    dplyr::select(flipper_length_mm, body_mass_g), 
  centers = k)

# extract cluster assignment per observation
Cluster = as.factor(penguins_k$cluster)

# extract cluster centers
ctrs <- as.data.frame(penguins_k$centers) %>% 
  mutate(Cluster = factor(1:k))

# plot points with voronoi diagram showing nearest centroid
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g, color = Cluster)) +
  geom_point() + 
  legend_pos +
  geom_voronoi(
    data = ctrs, aes(fill=Cluster), color = NA, alpha=0.5, outline = box) + 
  geom_point(
    data = ctrs, pch=23, cex=2, fill="black")
```

### Task
Show the Voronoi diagram for fewer (k=2) and more (k=8) clusters to see how assignment to cluster centroids work.

## Hierarchical Clustering

```{r}
# load dune dataset from package vegan
data("dune")

# show documentation on dataset if interactive
if (interactive())
  help(dune)
```

### Question
What are the rows and columns composed of in the `dune` data frame?

**Answer::** The `dune` data frame are observations of 30 species at 20 sites. The columns represent names of the species, abbreviated to 4+4 letters where the first four letters of the species and the last four are BLANK. 

### Calculate Ecological Distances on `sites`
```{r}
sites <- tribble(
  ~site, ~sp1, ~sp2, ~sp3,
    "A",    1,    1,    0,
    "B",    5,    5,    0,
    "C",    0,    0,    1) %>% 
  column_to_rownames("site")
sites
```

```{r}
sites_manhattan <- vegdist(sites, method="manhattan")
sites_manhattan
```


```{r}
sites_euclidean <- vegdist(sites, method="euclidean")
sites_euclidean
```


```{r}
sites_bray <- vegdist(sites, method="bray")
sites_bray
```


### Agglomerative hierarchical clustering on `dune`

```{r}
# Dissimilarity matrix
d <- vegdist(dune, method="bray")
dim(d)
```


```{r}
as.matrix(d)[1:5, 1:5]
```

```{r}
# Hierarchical clustering using Complete Linkage
hc1 <- hclust(d, method = "complete" )

# Dendrogram plot of hc1
plot(hc1, cex = 0.6, hang = -1)
```


```{r}
# Compute agglomerative clustering with agnes
hc2 <- agnes(dune, method = "complete")

# Agglomerative coefficient
hc2$ac
```


```{r}
# Dendrogram plot of hc2
plot(hc2, which.plot = 2)
```

```{r}
# methods to assess
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

# function to compute coefficient
ac <- function(x) {
  agnes(dune, method = x)$ac
}

# get agglomerative coefficient for each linkage method
purrr::map_dbl(m, ac)
```


```{r}
# Compute ward linkage clustering with agnes
hc3 <- agnes(dune, method = "ward")

# Agglomerative coefficient
hc3$ac
```


```{r}
# Dendrogram plot of hc3
plot(hc3, which.plot = 2)
```


### Divisive hierarchical clustering on `dune`

```{r}
# compute divisive hierarchical clustering
hc4 <- diana(dune)

# Divise coefficient; amount of clustering structure found
hc4$dc
```

### Determining optimal clusters

```{r}
# Plot cluster results
p1 <- fviz_nbclust(dune, FUN = hcut, method = "wss",  k.max = 10) +
  ggtitle("(A) Elbow method")

p2 <- fviz_nbclust(dune, FUN = hcut, method = "silhouette", k.max = 10) +
  ggtitle("(B) Silhouette method")

p3 <- fviz_nbclust(dune, FUN = hcut, method = "gap_stat", k.max = 10) +
  ggtitle("(C) Gap statistic")

# Display plots side by side
gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
```


### Working with dendrograms
```{r}
# Construct dendorgram for the Ames housing example
hc5 <- hclust(d, method = "ward.D2" )
dend_plot <- fviz_dend(hc5)
```


```{r}
dend_data <- attr(dend_plot, "dendrogram")
dend_cuts <- cut(dend_data, h = 8)
fviz_dend(dend_cuts$lower[[2]])
```


```{r}
# Ward's method
hc5 <- hclust(d, method = "ward.D2" )

# Cut tree into 4 groups
k = 4
sub_grp <- cutree(hc5, k = k)

# Number of members in each cluster
table(sub_grp)
```


```{r}
# Plot full dendogram
fviz_dend(
  hc5,
  k = k,
  horiz = TRUE,
  rect = TRUE,
  rect_fill = TRUE,
  rect_border = "jco",
  k_colors = "jco")
```

